---
title: "Ithaphen - Analysis and generating paper plots"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 6
    mathjax: null
    css: doc.css
params:
  input_path: ./data
  output_path: ./results
  plot_path: ./plots
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE)
```

<!-- Import required libraries -->

```{r warning=FALSE, message=FALSE, echo=FALSE}
# for reproducible results
set.seed(1989)
options(box.path = getwd())
#load libraries
box::use(
  dplyr[d_filter = filter, d_select = select, ...],
  # for unite
  tidyr[...],
  # for enframe  
  tibble[...],
  readr[...],
  # for map_chr
  purrr[...],
  stringr[...],
  knitr[...],
  # For pretty-printed tables
  pander[...],
  readxl[...], 
  xlsx[write.xlsx],
  DBI[...],
  odbc[...],
  RMariaDB[...],
  ggplot2[...],
  ggrepel[...],
  cowplot[...],
  patchwork[...],
  RecordLinkage[...],
  countrycode[...],
  ggupset[...],
  ontologyIndex[...],
  tidytext[...],
  corrplot[...],
  skimr[...],
  # For pretty-printed tables
  kableExtra[...],
  source/aux_plotting[...]
  )

gg_colors = gg_color_hue(10)[c(1, 2, 5, 8, 9, 10)]
paper_colors = c("#959796", # grey
                 "#cd5644", #rooftile
                 "#da9537", #orange
                 "#756161", #brown
                 "#8a8f3e", #vomitgreen
                 "#7263d3", #lavender
                 "#5ba5db", #blue
                 "#d544a0", #magenta
                 "#5db852", #green
                 "#d08cca", #pink
                 "#ac53bc", #purple/magenta
                 "#4fb8a1" #mint
                 )
theme_set(theme_cowplot(font_size = 12))
```

```{r demographics, message=FALSE, echo=FALSE}
pd_1a  = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1A") %>% 
  mutate(ancestor_name = ifelse(str_detect(ancestor_name, "^Greater Middle Eastern"), 
                                # write this in the legend cause it doesn't fit
                                #"Greater Middle Eastern\n(Middle Eastern, North African\nor Persian)", ancestor_name),
                                "Greater Middle Eastern", ancestor_name),
         ancestor_name = ifelse(str_detect(ancestor_name, " or\\s"), 
                                str_replace_all(ancestor_name, "\\sor\\s", "/"), 
                                ancestor_name))

f_1a  = ggplot(pd_1a ,
                  aes(y = reorder(ancestor_name, percentage), x = percentage)) + 
  geom_bar(stat = "identity", colour = "#959796", fill = "#959796") +
  geom_text(aes(x = percentage,
                label = ifelse(count < 300, "", count)),
            hjust = 1.1,
            size = 4, col = "white") +
  scale_x_continuous(limits = c(0, ceiling(max(pd_1a$percentage) / 10 ) * 10),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  labs(y="", x = "% of total cases", fill="") + 
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none", 
        #aspect.ratio = 2,
        panel.spacing = unit(1.5, "lines"), 
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

pd_1b_bottom  = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1B Bottom")

f_1b_bottom = ggplot(pd_1b_bottom, aes(y=sex)) + 
  geom_bar(aes(x = percentage), stat = "identity", 
           colour = "#959796", fill = "#959796", width = 0.75) +
  geom_text(aes(x = percentage,
                label = ifelse(percentage < 5, "", count)),
            hjust = 1.1, size=4, colour="white") +
  scale_x_continuous(limits = c(0, ceiling(max(pd_1b_bottom$percentage)/10) * 10),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  labs(y = "Sex",
       x = "% of total cases") +
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none", 
        panel.spacing = unit(1.5, "lines"),
        axis.line.y = element_blank(), 
        axis.ticks.y = element_blank())

pd_1b_top = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1B Top") %>% 
  mutate(age_group = factor(age_group, levels = c("Prenatal", "Under 5", "5-19", "Above 19", "Unknown")))

f_1b_top = ggplot(pd_1b_top, aes(y = age_group, x = percentage)) +
  geom_bar(stat = "identity", colour = "#959796", fill = "#959796", width = 0.75) +
  geom_text(aes(x = percentage,
                label = ifelse(percentage < 1, "", count)),
            hjust = 1.1, size = 4, colour = "white") +
  scale_x_continuous(limits = c(0, ceiling(max(pd_1b_top$percentage) / 10 ) * 10),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  labs(y = "Age",
       x = "% of total cases") +
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none",
        panel.spacing = unit(1.5, "lines"),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank()) 

pd_1b = pd_1b_bottom %>% mutate(group = "Sex") %>%
  rename(y = sex) %>%
  bind_rows(pd_1b_top %>% mutate(group = "Age") %>%
              rename(y = age_group)) %>% 
  mutate(y = factor(y, levels = c("Female", "Male", "Prenatal", "Under 5", "5-19", "Above 19", "Unknown")))
f_1b = ggplot(pd_1b, aes(y = y, x = percentage)) +
  facet_wrap(~group, nrow =2, scales="free_y") +
  geom_bar(stat = "identity",
           colour = "#959796", fill = "#959796",
           width = 0.75
           ) +
  geom_text(aes(x = percentage,
                label = ifelse(percentage < 1, "", count)),
            hjust = 1.1, size = 4, colour = "white") +
  scale_x_continuous(limits = c(0, ceiling(max(pd_1b$percentage) / 10 ) * 10),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  labs(y = "",
       x = "% of total cases") +
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none",
        strip.placement = "outside",
        strip.background = element_rect(fill = "white", linewidth = 1),
        panel.spacing = unit(1.5, "lines"),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank()) 


```

```{r per-genotype, message=FALSE, echo=FALSE}

pd_1c  = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1C") %>%
  mutate(current = factor(current, levels = c("None", "Homozygous", "Heterozygous",  "Compound heterozygous")), 
         current = recode_factor(current, !!!setNames(c("None", "\nHomozygous", "Heterozygous",  "\nCompound\nheterozygous"), 
                                                 c("None", "Homozygous", "Heterozygous",  "Compound heterozygous"))),
         complement = factor(complement, levels = c("Compound heterozygous", "Homozygous", "Heterozygous", "None")))

f_1c  = ggplot(pd_1c, aes(x = current, y=percentage, fill = complement)) +
  facet_wrap(~locus, ncol = 2, strip.position = "top") +
  geom_bar(stat = "identity", width = 0.8) +
  geom_label_repel(aes(label=count, fill=complement), 
                   position= position_stack(vjust= 0.5), 
                   direction = "both",
                   segment.color="grey15", box.padding = unit(0.3, 'lines'),
                   min.segment.length = 0.1, label.padding = 0.2,
                   size=4, colour="grey15",
                   show.legend = F, seed = 378) +
  scale_y_continuous(# this is so that 0 is at the intersection of x axis with y
                    expand = expansion(mult = c(0, 0)),
                    limits = c(0,100)) +
  # Adding line break to compound heterozygous
  scale_x_discrete(labels = str_replace(levels(pd_1c$current)[str_detect(levels(pd_1c$current), "ygous$")], "\\s", "\n")) + 
  scale_fill_manual(values= setNames(paper_colors[c(1, 3, 2, 4)], c("None", "Heterozygous", "Homozygous", "Compound heterozygous"))) +
  guides(fill=guide_legend(nrow = 2, byrow = T)) + 
  labs(x="", y = "% of total cases on locus", fill="Other globin locus") + 
  theme_cowplot(font_size = 12) +
  theme(legend.position="bottom", 
        #aspect.ratio = 0.25,
        strip.placement = "outside",
        panel.spacing = unit(1.5, "lines"),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

```{r per-param, message=FALSE, echo=FALSE}
pd_1d  = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1D")

f_1d  = ggplot(pd_1d,
             aes(y=reorder(ancestor_name, percentage), x=percentage, fill = ancestor_name)) +
  geom_bar(stat= "identity", width = 0.75) +
  geom_text(aes(x= percentage, 
                label = ifelse(count < 720, "", count)),
            hjust = 1.1, size=4, colour="white") +
  scale_fill_manual(values=setNames(colfunc(length(levels(as.factor(pd_1d[['ancestor_name']])))),
                                    pd_1d %>%
                                       group_by(ancestor_name) %>%
                                       summarise(count = n()) %>%
                                       arrange(count) %>% .[['ancestor_name']] )) +
  scale_x_continuous(limits = c(0, 100),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  labs(y="", x = "% of total cases", fill="Category") + 
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none", 
        panel.spacing = unit(1.5, "lines"),
        axis.text.y = element_text(angle = 0),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

pd_1e  = read_xlsx(file.path(params$input_path, "plot-data.xlsx"),
                   sheet = "Figure 1E")
  
f_1e  = ggplot(pd_1e %>% d_filter(count > 400) %>% group_by(ancestor_name), 
             aes(y=reorder_within(parameter_name, percentage, ancestor_name), x=percentage, fill = ancestor_name)) +
  geom_bar(stat= "identity") +
  geom_text(aes(x= percentage, 
                label = ifelse(count < 800, "", count)),
            hjust = 1.1, size=4, colour="white") +
  scale_y_reordered() +
  scale_x_continuous(limits = c(0,100),
                     # this is so that 0 is at the intersection of x axis with y
                     expand = expansion(mult = c(0, 0))) +
  scale_fill_manual(values=setNames(colfunc(length(levels(as.factor(pd_1e[['ancestor_name']])))),
                                    pd_1e %>%
                                       group_by(ancestor_name) %>%
                                       summarise(count = n()) %>%
                                       arrange(count) %>% .[['ancestor_name']] )) +
  labs(y = "", #y="Parameter",
       x = "% of total cases", fill="Category") + 
  theme_cowplot(font_size = 12) + 
  theme(legend.position="none", 
        panel.spacing = unit(1.5, "lines"),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

```

```{r figure-1nS1, message=FALSE, echo=FALSE}

f_1 = f_1a + f_1b  + f_1c  + f_1d + f_1e + guide_area() +
  plot_layout(guides="collect", 
              heights = c(1,1.25,1.5, 0.3),
              widths = c(1, 1), 
              design ="AB\nCC\nEF\nGG") +
  plot_annotation(tag_levels = "A") 
ggsave(file.path(params$plot_path, "Figure-1.pdf"), 
       f_1,  
       bg = "white",
       height = 29.52,
       width =  21.01,
       units = "cm",
       device = grDevices::cairo_pdf,
       dpi="retina")
```

# Demonstration of IthaPhen utility - Example 1

## Dataset

```{r load-data, echo = FALSE, results='asis'}
box::use(marginaleffects[...], 
         sjPlot[...], 
         modelsummary[...],
         caret[...])

ds = read_xlsx(file.path(params$input_path, 
                         "db-thal-n-HPFH-heterozygote-cases.xlsx"),
               sheet = "cases")

ds %>% 
  group_by(outcome) %>% 
  summarise(N = n()) %>% 
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Dataset used in stepwise logistic regression") %>% 
  kable_styling(full_width = T) %>%
  print()

ds %>%
  d_select(pid, `Hb F`, `Hb A2`, Hb, MCV, MCH, RBC) %>%
  pivot_longer(-pid, names_to = "parameter", values_to = "value") %>%
  group_by(parameter) %>%
  d_filter(is.na(value)) %>% 
  summarise(missing_obs = n()) %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Missing observation per phenotypic parameter") %>% 
  kable_styling(full_width = T) %>%
  print()
```

## Stepwise logistic regression

Basic assumptions that must be met for *logistic* regression include:
 1. independence of errors
 2. linearity in the logit for continuous variables
 3. absence of multicollinearity: Multicollinearity occurs when two or more independent variables have a high correlation with one another in a regression model, which makes it difficult to determine the individual effect of each independent variable on the dependent variable.
 4. lack of strongly influential outliers.
 
In a regression model, consider including the interaction between 2 variables when:

 1. They have large main effects: Variables that have a large influence on the 
 outcome are more likely to have a statistically significant interaction with other factors that influence this outcome.
 2. The effect of one independent variable changes for various subgroups of the other.
 3. The interaction has been proven in previous studies.
 

```{r stepwise-regression, message=FALSE, echo = FALSE, warning=FALSE,results='asis'}
clean_df0 = ds %>% 
  d_select(-c(pid, state, family_id)) %>%
  mutate(across(where(is.character), ~str_replace(., "\\s", '_')),
         across(where(is.character), ~str_replace(., "\\s", '_')),
         across(where(is.character), ~as.factor(.))) %>% 
  mutate(imputed_age = ifelse(is.na(formatted_age), 
                              round(median(na.omit(formatted_age))), formatted_age))

contr = contrasts(clean_df0$outcome)
clean_df0 = clean_df0 %>%
  mutate(outcome = recode(outcome,
                          !!!setNames(contr %>% as_tibble() %>% .[[1]], 
                                      contr %>% rownames())))

r = names(clean_df0 %>% d_select(-outcome)) %>% map(function(x) {
  df= clean_df0 %>% d_select(outcome, !!!x)
  m = glm(outcome ~ ., family = binomial(link="logit"), data = df)
  if(coef(summary(m))[2,4] < 0.05){
    x
  }
  }) %>% unlist()

# this is with classes for age and parameters found important are Hb F, MCV, age, RBC
# num obs = 86
clean_df = clean_df0 %>%
  d_select(any_of(r), outcome) %>%
  drop_na(everything())
    
null_model = glm(outcome ~ 1,
                     family = binomial(link = "logit"),
                     data = clean_df)
full_model = glm(outcome ~ .,
                     family = binomial(link = "logit"),
                     data = clean_df)

# Perform stepwise regression
stepwise_r = MASS::stepAIC(null_model, # start with a model containing no variables
                direction = 'both', # run forward and backward selection
                scope = list(upper = full_model, # the maximum to consider is a model with all variables
                             lower = null_model), # the minimum to consider is a model with no variables
                trace = 0) # do not show the step-by-step process of model selection

stepwise_r$anova %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "ANOVA table for final model from stepwise regression") %>% 
  kable_styling(full_width = T) %>%
  print()
```

## Final additive logistic regression model

Having identified significant predictors of δβ-thalassemia with stepwise 
regression analysis, we build an additive logistic regression model with those 
predictors using all the data with non missing pairwise observations.

```{r additive-logististic-regression-1, message=FALSE, echo = FALSE, results='asis'}
clean_df = clean_df0 %>%
  d_select(any_of(predictors(stepwise_r)), outcome) %>%
  drop_na(everything())

clean_df %>% 
  # recoding the outcome from binary back to factor before the LR analysis
  mutate(outcome = recode(outcome,
                          # ensuring american spelling for thalassemia at the same
                          # time
                          !!!setNames(contr %>% rownames(),
                                      contr %>% as_tibble() %>% .[[1]]))) %>%
  group_by(outcome) %>% 
  summarise(N = n()) %>% 
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Dataset used in final logistic regression model") %>% 
  kable_styling(full_width = T) %>%
  print()

additive_model = glm(outcome ~ .,
                     family = binomial(link = "logit"),
                     data = clean_df)

# McFadden's pseudo R^2 can be interpreted like R^2, but don't expect it to be as big. And values from 0.2-0.4 indicate (in McFadden's words) excellent model fit. 
# McFadden's R^2 is defined as 1−LLmod/LL0, where LLmod  is the log likelihood value for the fitted model and LL0 is the log likelihood for the null model which includes only an intercept as predictor (so that every individual is predicted the same probability of 'success').
# For a logistic regression model the log likelihood value is always negative (because the likelihood contribution from each observation is a probability between 0 and 1). If your model doesn't really predict the outcome better than the null model, LLmod will not be much larger than LL0, and so LLmod/LL0 ~=1, and McFadden's pseudo-R^2 is close to 0 (your model has no predictive value).
pscl::pR2(additive_model) %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "pseudo-R2 measures for additive model") %>% 
  kable_styling(full_width = T) %>%
  print()
```

### Checks for linearity and homoscedascity


```{r additive-logististic-regression-2, message=FALSE, echo = FALSE, warning=FALSE, results='asis'}

# Check for linearity and homoscedascity (equal/similar variances across groups 
# being compare) assumption violation
# When the residuals have mean zero for every value of the fitted values and of
# the predictors all that relevant variables and interactions are included in 
# the model, and the functional form of the relationship between the predictors 
# and the outcome is correct.
# Any association between the residuals and fitted values or predictors implies
# unobserved confounding (i.e., endogeneity), and no causal interpretation can
# be drawn from the model.
# Cone-shape is indicative of heterscedascity
p = ggplot(tibble(residuals = additive_model$residuals,
              fitted_values = additive_model$fitted.values) %>%
             bind_cols(clean_df), 
       aes(fitted_values, residuals) ) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") + 
  geom_smooth()
print(p)

# If the residuals mean deviates from zero, there is evidence for non-linearity 
p = ggplot(tibble(residuals = additive_model$residuals,
              fitted_values = additive_model$fitted.values) %>%
             bind_cols(clean_df), 
       aes(log(`Hb F`), residuals) ) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") + 
  geom_smooth()
print(p)

p = ggplot(tibble(residuals = additive_model$residuals,
              fitted_values = additive_model$fitted.values) %>%
             bind_cols(clean_df), 
       aes(log(MCV), residuals) ) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") + 
  geom_smooth()
print(p)

# variance inflation factor (VIF) to detect multicollinearity
# How much of one's predictor's variance is explained by the other predictors
# It is predicted by taking a variable and regressing it against every other variable
# Perfectly uncorrelated predictors have VIFs of 1,
# and perfectly correlated predictors have VIFs of infinity.
# Although correlation matrix and scatter plots can also be used to find multicollinearity, their findings only show the bivariate relationship between the independent variables. VIF is preferred as it can show the correlation of a variable with a group of other variables. 
car::vif(additive_model) %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Variance inflation factor") %>% 
  kable_styling(full_width = T) %>%
  print()
```

```{r comparing-models, message=FALSE, echo = FALSE, warning=FALSE, results='asis'}
cross_model = glm(outcome ~ `Hb F` * MCV,
                     family = binomial(link = "logit"),
                     data = clean_df)

modelsummary(list("Stepwise" = stepwise_r, "Additive" = additive_model, "Cross" = cross_model), 
             exponentiate = F, 
             stars=T, 
             estimate = "{estimate} ({std.error}){stars}",
             output="DT",
             extensions = "FixedColumns",
             options= list(pageLength = 50, 
                           scrollX = TRUE, 
                           fixedColumns = list(leftColumns = 1)),
             statistic = "{p.value} [{conf.low}, {conf.high}]")


# Odds ratio plot for additive model
# Can not use axis.lim option to provide more space to the right of the vline=1
# so will not be using this plot. Instead using the plot data generated by it
# mainly because it already has formatted the labels to plot it and tweak the axis
p0 = plot_model(additive_model, 
                vline.color = "red",
                show.values = T, 
                sort.est = T) 
p0$data %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Odds ratio") %>% 
  kable_styling(full_width = T) %>%
  print()

clean_df = clean_df %>% 
  # recoding the outcome from binary back to factor before the LR analysis
  mutate(outcome = recode(outcome,
                          # ensuring american spelling for thalassemia at the same
                          # time
                          !!!setNames(str_replace(contr %>% rownames(), 
                                                  "thalassaemia", "thalassemia"),
                                      contr %>% as_tibble() %>% .[[1]])))
pd_2a = clean_df %>%
  pivot_longer(-outcome, names_to = "index", values_to = "value") %>%
  rename(Condition = outcome) %>%
  mutate(index = ifelse(index == "Hb F", "Hb F (%)", "MCV (fL)"))
f_2a = ggplot(pd_2a, aes(x=value, colour = Condition, fill = Condition)) +
  geom_histogram(binwidth= 2, position = position_identity(), 
                 # use boundary 0 to set the left boundary of the first bin to
                 # 0 and not 0.5. The counts in the bins are based on condition
                 # value > bin & value <= bin + bin_width
                 boundary = 0,
                 alpha=0.5) +
  # the numbers are correct only if using facet_grid, facet_wrap has incorrect
  # y-axis behaviour
  facet_grid(index~.) + #, scales = "free", nrow = 2) +
  scale_fill_manual(values=c("#c2dc75", "#56dee6")) +
  scale_colour_manual(values=c("#c2dc75", "#56dee6")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  labs(y = "Number of cases", x = "") + 
  theme_cowplot(font_size = 12) +
  theme(legend.position="bottom", 
        panel.spacing = unit(1.5, "lines")) 

f_2b = ggplot(p0$data, aes(y=term, x=estimate, 
                           #color = group,
                           label = p.label,
                           xmin=conf.low, xmax=conf.high )) +
  geom_pointrange(size = 0.9, shape = 20, color = "grey20", 
                  position = position_dodge2(width = 0.5, preserve = "total")) +
  geom_text(nudge_y = 0.1, nudge_x = 0, size =  4) +
  geom_vline(aes(xintercept = 1), color = "red", linetype = "solid") +
  scale_x_continuous(expand = expansion(mult = c(0, 0)),
                     limits = c(0.7, 1.1)) + 
  #scale_color_manual(values=setNames(c("red", "blue"), c("neg", "pos"))) +
  labs(#subtitle = str_c(colnames(contr), " ~ ", str_c(predictors(additive_model), collapse=" + "), sep=""), 
       x = "Odds Ratio", y = "") +
  theme_cowplot(font_size = 12) + 
  theme(panel.grid.major.x = element_line(colour = "grey92"), 
        panel.grid.minor.x = element_line(colour = "grey92", linewidth = 0.25), 
        panel.grid.major.y = element_line(colour = "grey92"), 
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") 
  
```

# Demonstration of IthaPhen utility - Example 2

## LR computation

```{r LR-computation, echo=FALSE, message=FALSE, results='asis'}
stats = clean_df %>% 
  summarise(across(c(`Hb F`, "MCV"),
                   .fns = list(Mean = mean, SD = sd, Q = quantile),
                   .names = "{col}_{fn}")) %>% 
  mutate(q = str_c("q_", seq(0, 1, 0.25)), 
         outcome = "any") %>% 
  pivot_longer(-c(outcome, q), names_to = "tmp") %>% 
  separate(tmp, into = c("param", "metric"), sep="_") %>% 
  mutate(metric = ifelse(metric == "Q", q, metric)) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  group_by(outcome, param) %>% fill(everything(), .direction ="updown") %>%
  d_select(-q) %>% 
  distinct() %>%
  mutate(r = round(SD / 10, digits = 1), 
         range = list(seq(round(q_0.25, digits = 1), 
                          round(q_0.75, digits=1), 
                          by = r)), 
         # range of interest
         roi = case_when(param == "Hb F" ~ list(c(seq(10, 25, 5), 100)), 
                         param == "MCV" ~ list(c(seq(70, 80, 5), 100)))) %>%
  ungroup()  

lr = stats %>%
  d_filter(outcome == "any") %>%
  unnest(roi) %>% 
  group_by(param) %>% 
  mutate(r = pmap(.l = list(lag(roi), roi), 
                  .f = function(x, y){
                    list(x, y)
                  })) %>% 
  d_select(param, r) %>%
  rowwise() %>% 
  group_split() %>%
  map(function(x){
    p = x$param
    r = unlist(x$r)
    tmp  = str_c(p, str_c(sort(na.omit(r)), collapse = "_"), sep="_")
    clean_df %>% 
      rename(focal := {{p}} ) %>%
      mutate(wr = case_when((is.na(r[1]) & focal <= r[2]) ~ 1, 
                            (is.na(r[1]) & focal > r[2]) ~ 0, 
                            (!is.na(r[1]) & focal <= r[1]) ~ 0, 
                            (!is.na(r[1]) & focal > r[2]) ~ 0,
                            (focal > r[1] & focal <= r[2]) ~ 1)) %>%
      group_by(outcome) %>%
      summarise(wr = sum(wr)) %>%
      rename({{tmp}} := wr)
      }) %>%
  Reduce(function(df1, df2) bind_cols(df1, df2 %>% ungroup() %>% d_select(-outcome)), .) %>%
  pivot_longer(-outcome, names_to = "th") %>%
  pivot_wider(names_from = "outcome", values_from = "value") %>%
  separate(th, into = c("param", "th_1", "th_2"), sep="_", fill="right") %>%
  group_by(param) %>%
  # the first range is starting from 0 up to th_1 based on the comparisons above 
  # so formatting the dataframe accordingly
  mutate(th_2 = ifelse(is.na(th_2), th_1, th_2),
         th_1 = ifelse(th_1==th_2, 0, th_1),  
         across(c(`δβ-thalassemia`, `deletional-HPFH`), \(x) x/sum(x), .names = "{.col}_prob"),
         LR = `δβ-thalassemia_prob` / `deletional-HPFH_prob`,
         # calculate the standard error of the LR
         LR_se  = sqrt(1/`δβ-thalassemia` + 1/`deletional-HPFH` -1/sum(`δβ-thalassemia`) -1/sum(`deletional-HPFH`)), 
         across(LR,
                function(x){
                  get(glue::glue("{cur_column()}_se"))^2 
                  }, .names = "{.col}_variance"),
         # nifty little code to calculate lower CI on both LR + and LR - because
         # getting the value of the se for the specific LR with glue that builds the name 
         # from the current column and get that gets the column value
         across(LR, function(x){exp(log(x) - 1.96 * get(glue::glue("{cur_column()}_se")))}, .names = "{.col}_lCI"),
         across(LR, function(x){exp(log(x) + 1.96 * get(glue::glue("{cur_column()}_se")))}, .names = "{.col}_uCI")) %>%
  d_select(param, th_1, th_2, 
           contains("thalassemia"), contains("deletional"), 
           starts_with("LR"))

```

```{r echo=FALSE, message=FALSE, results='asis'}
r = lr %>% 
  mutate(LR_formatted = str_c(round(LR, 3), " (", round(LR_lCI, 3), ", ", round(LR_uCI, 3), ")", sep="")) %>%
  d_select(1:7, LR_formatted) %>%
  mutate_if(is.numeric, round, 3)
  

r %>%
  as.data.frame() %>%
  kable(align = "c", 
        caption = "Likelihood ratios") %>% 
  kable_styling(full_width = T) %>%
  print()

write.xlsx(r %>% as.data.frame(),
           file.path(params$output_path, 'likelihood-ratio-table.xlsx'),
           sheetName = "LR", append=F)

```

```{r figure-2, echo=FALSE, message=FALSE}
pd_2c = lr %>% 
  mutate(across(c(th_1, th_2), as.numeric), 
         r = str_c(th_1, th_2, sep="-"), 
         param = recode(param, !!!setNames(c("Hb F (%)", "MCV (fL)"), c("Hb F", "MCV")))) %>%
  arrange(param)
  
f_2c = ggplot(pd_2c, aes(r, LR, ymin=LR_lCI, ymax=LR_uCI, group = r)) +
  geom_pointrange(size = 0.5, shape = 20, color = "grey20", 
                  position = position_dodge2(width = 0.5, preserve = "total")) +
  facet_wrap(param~., scales="free") +
  guides(colour = guide_legend(title = "")) +
  geom_hline(aes(yintercept = 1), linetype = "solid", color = "red") +
  geom_text(aes(0.25, 1, label = "LR = 1", vjust = -0.25, hjust = -0.25),
            size = 3, color = "red", check_overlap = T) +
  scale_y_continuous(trans='log2') + 
  theme_cowplot(font_size = 12) + 
  labs(y ="Likelihood Ratio", x="") 


f_2 = f_2a + f_2b + f_2c + guide_area() +
  plot_layout(guides = "collect",
              widths = c(1, 1),
              heights = c(0.75, 1.25, 0.15),
              design = "AB\nCC\nDD") +
  plot_annotation(tag_levels = "A")


ggsave(file.path(params$plot_path, "Figure-2.pdf"), 
       f_2,  
       bg = "white",
       width = 21.01,
       height = 29.62,
       units = "cm",
       device = grDevices::cairo_pdf,
       dpi="retina")
```

